---
title: "Thesis Preliminary Analysis"
author: "Alice Tivarovsky"
date: "1/13/2020"
output: html_document
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE)

library(tidyverse)
library(foreign)
library(ggridges)
library(viridis)
library(tableone)
library(survey)
library(stats)


options(
  ggplot2.continuous.colour = "viridis",
  ggplot2.continuous.fill = "viridis"
)
scale_colour_discrete = scale_colour_viridis_d
scale_fill_discrete = scale_fill_viridis_d
theme_set(theme_minimal() + theme(legend.position = "bottom"))

```

## Data Import and Tidying

Reading in cycles 2015-2016 and 2013-2014, datasets: demographics (DEMO), diet behavior and nutrition (DBQ), Phthalates and Plasticizers Metabolites - Urine (PHTHTE), Personal Care and Consumer Product Chemicals and Metabolites (EPHPP) in the 2015-2016 and 2013-2014 cycles, and Environmental Phenols & Parabens (EPH_G) 2011-2012 cycles. 

```{r}
# reading in 2015 - 2016 data
demo_15_16 = read.xport("../data/2015_2016_DEMO_I.XPT.txt")
dbq_15_16 = read.xport("../data/2015_2016_DBQ_I.XPT.txt")
phthalate_15_16 = read.xport("../data/2015_2016_PHTHTE_I.XPT.txt")
phenol_15_16 = read.xport("../data/2015_2016_EPHPP_I.XPT.txt")

# reading in 2013 - 2014 demographics data
demo_13_14 = read.xport("../data/2013_2014_DEMO_H.XPT.txt")
dbq_13_14 = read.xport("../data/2013_2014_DBQ_H.XPT.txt")
phthalate_13_14 = read.xport("../data/2013_2014_PHTHTE_H.XPT.txt")
phenol_13_14 = read.xport("../data/2013_2014_EPHPP_H.XPT.txt")

```
 

Merging everything into one dataframe: 

```{r}
data_15_16 = 
  full_join(demo_15_16, dbq_15_16, by = "SEQN") %>%
  full_join(phthalate_15_16, by = "SEQN") %>% 
  full_join(phenol_15_16, by = "SEQN") %>% 
  mutate(cycle = "2015_2016")

data_13_14 = 
  full_join(demo_13_14, dbq_13_14, by = "SEQN") %>%
  full_join(phthalate_13_14, by = "SEQN") %>% 
  full_join(phenol_13_14, by = "SEQN") %>% 
  mutate(cycle = "2013_2014")
  
all_data = 
  bind_rows(data_13_14, data_15_16) 
```


Select variables of interest: 

```{r}
all_data = 
  all_data %>% 
  select(SEQN, cycle, RIAGENDR, RIDAGEYR, RIDRETH1, RIDRETH3, DMDEDUC2, DMDEDUC3, DMDMARTL, INDHHIN2, INDFMIN2, WTMEC2YR, SDMVPSU, SDMVSTRA, URXBP3, URXBPH, URXBPF, URXBPS, URXTRS, URXUCR.x, URXCNP, URXCOP, URXECP, URXHIBP, URXMBP, URXMC1, URXMCOH, URXMEP, URXMHBP, URXMHH, URXMHNC, URXMHP, URXMIB, URXMNP, URXMOH, URXMZP, DBQ700, DBD895, DBD900, DBD905, DBD910) 

```
We start with 20,146 observations before any exclusion criteria are applied. 

Restrict to women 18 - 44 with biomarker data.  
```{r}
final_data = 
  all_data %>% 
  filter(RIAGENDR == 2, RIDAGEYR %in% (18:44)) %>% 
  drop_na(URXBP3)

#check
min(final_data$RIDAGEYR)

```

Next, we recode the education variables because education for 18-20 year olds is coded differently from those > 20 years old. First we get the range to understand the scope of recode. 

```{r} 
range(all_data$DMDEDUC2, na.rm = TRUE)

all_data_1 = 
  final_data %>% 
  filter(RIDAGEYR %in% (18:19)) %>% 
    mutate(DMDEDUC2 = as.numeric(DMDEDUC2), 
          DMDEDUC3 = as.numeric(DMDEDUC3)) %>% 
     mutate(DMDEDUC2 = 
            ifelse(DMDEDUC3 < 9, "1", 
            ifelse(DMDEDUC3 %in% (9:11), "2",
            ifelse(DMDEDUC3 %in% (12:14), "3", "4")))) %>% 
  mutate(DMDEDUC2 = as.numeric(DMDEDUC2))
  

all_data_2 = 
  final_data %>% 
  filter(!RIDAGEYR %in% (18:19))

final_data = bind_rows(all_data_1, all_data_2)

```


Now, we recode marital status for 18 and 19-year olds, who were not asked the question. NA's are replaced with 5 (Never Married). 
```{r}
final_data = 
  final_data %>% 
  mutate(DMDMARTL = replace_na(DMDMARTL, 5))

```

We note that we have 2 income variables, annual household income (INDHHIN2) and annual family income (INDFMIN2). They have the same coding structure, and are likely highly correlated. We will remove one of them for further analysis, but first we confirm the correlation. Running the correlation below, r = 0.9720947, confirming correlation. We proceed with only family income. 

```{r income correlation}

final_data %>% 
  select(INDHHIN2, INDFMIN2) %>% 
  cor(use = "complete.obs", method = "pearson")
```


Finally, we clean up the variable names and types and remove NA vectors (URXHIBP, URXMCOH, URXMHBP). We omit household income, per the correlation above, and the first ethnicity variable (ridreth1), which is less information than ridreth3. 

```{r}
# Fixing column labels
final_data = 
  final_data %>% 
  janitor::clean_names() %>% 
  mutate(ridageyr = as.integer(ridageyr), 
         indfmin2 = as.factor(indfmin2), 
         ridreth3 = as.factor(ridreth3), 
         dmdeduc2 = as.factor(dmdeduc2),
         ) %>% 
  select(-riagendr, -ridreth1, -indhhin2, -dmdeduc3, -urxhibp, -urxmcoh, -urxmhbp)

save(final_data, file = "final_data.Rdata")
```


## Exploratory data analysis

### Distributions

Independent variable histograms: 
```{r independent variable histograms}

# Age histogram
final_data %>% 
  ggplot(aes(x = ridageyr)) + 
  geom_histogram(position = "dodge", binwidth = 1, color = "darkblue", fill = "lightblue") 

# Education histogram
final_data %>% 
  ggplot(aes(x = dmdeduc2)) +
  geom_histogram(position = "dodge", binwidth = 1, color = "darkgreen", fill = "lightgreen", stat = "count")

# Household Income Histogram
final_data %>% 
  ggplot(aes(x = indfmin2)) +
  geom_histogram(position = "dodge", binwidth = 1, color = "red", fill = "pink",
                 stat = "count")

```

Outcome variable histograms (to understand if log adjustment is necessary). First we look at phenols. The three phenol distributions indicate right-skewed data, suggesting the need to log-transform. 

```{r phenol distributions}

# urxbp3 Urinary Benzophenone-3 (ng/ml)
final_data %>% 
  ggplot(aes(x = urxbp3)) + 
  geom_histogram(position = "dodge", binwidth = 1, color = "darkblue", fill = "lightblue") 

# urxbph Urinary Bisphenol A (ng/mL)
final_data %>% 
  ggplot(aes(x = urxbph)) + 
  geom_histogram(position = "dodge", binwidth = 1, color = "darkblue", fill = "lightblue") 

# urxbpf Urinary Bisphenol F (ng/mL)
final_data %>% 
  ggplot(aes(x = urxbpf)) + 
  geom_histogram(position = "dodge", binwidth = 1, color = "darkblue", fill = "lightblue") 


```

Next we examine the phthalate distributions. Similarly, the phthalates are right-skewed. 

```{r pthalate distributions}

# urxmbp	Mono-n-butyl phthalate (ng/mL)
final_data %>% 
  ggplot(aes(x = urxmbp)) + 
  geom_histogram(position = "dodge", binwidth = 1, color = "darkblue", fill = "lightblue") 

# urxmep Mono-ethyl phthalate (ng/mL)
final_data %>% 
  ggplot(aes(x = urxmep)) + 
  geom_histogram(position = "dodge", binwidth = 1, color = "darkblue", fill = "lightblue") 

# urxmib Mono-isobutyl phthalate (ng/mL)
final_data %>% 
  ggplot(aes(x = urxmib)) + 
  geom_histogram(position = "dodge", binwidth = 1, color = "darkblue", fill = "lightblue") 

```



### Table 1  

```{r tableone}

tab_one_data = 
final_data %>% 
  select("ridageyr", "ridreth3", "dmdeduc2", "dmdmartl", "indfmin2") %>% 
  rename("Age" = "ridageyr", "Ethnicity" = "ridreth3", "Education" = "dmdeduc2", "MaritalStatus" = "dmdmartl", "FamilyIncome" = "indfmin2") %>% 
  mutate(
    Ethnicity = recode(Ethnicity, "1" = "Mexican American", "2" = "Other Hispanic", "3" = "Non-Hispanic White", "4" = "Non-Hispanic Black", "6" = "Non-Hispanic Asian", "7" = "Other Race - Including Multi-Racial", .default = NULL, .missing = NULL), 
    Education = recode(Education, "1" = "Less than High School", "2" = "Less than High School", "3" = "High school graduate/GED or equivalent", "4" = "Some college or AA degree", "5" = "College graduate or above", "7" = "Refused", "9" = "Don't Know", .default = NULL, .missing = NULL), 
    MaritalStatus = recode(MaritalStatus, "1" = "Married", "2" = "Widowed", "3" = "Divorced", "4" = "Separated", "5" = "Never married", "6" = "Living with partner", .default = NULL, .missing = NULL))

#recoding income
tab_one_data =
  tab_one_data %>% 
  mutate(FamilyIncome = case_when(FamilyIncome %in% c("1", "2", "3", "4", "13") ~ "< $20,000", 
                                  FamilyIncome %in% c("5", "6", "7") ~ "$20,000 - $45,000",
                                  FamilyIncome %in% c("8", "9", "10") ~ "$45,000 - $75,000",
                                  FamilyIncome == 14 ~ "$75,000 to $100,000",
                                  FamilyIncome == 15 ~ "$100,000 and Over")
  ) %>% 
  mutate(FamilyIncome = as.factor(FamilyIncome)) %>% 
  drop_na(FamilyIncome)



my_vars = c("Age", "Ethnicity", "Education", "MaritalStatus", "FamilyIncome")

cat_vars = c("Ethnicity", "Education", "MaritalStatus", "FamilyIncome")

tab_one = CreateTableOne(vars = my_vars, data = tab_one_data, factorVars = cat_vars) 
kableone(tab_one)


```

## Modeling

Due to the complex sampling structure of NHANES, we need to account for survey weights. We use the survey package and subset our desired population (women aged 18 - 44). First we perform the same tidying steps we used for the exploratory analysis. We note that we cannot use the same dataset used in the exploratory analysis because application of the survey weights requires that we retain all observations and restrict the data using subsetting rather than filtering. 

```{r tidying for survey weights}
# assign variable types and remove unneeded variables 
model_data = 
  all_data %>% 
  janitor::clean_names() 

# recoding education
model_data_1 = 
  model_data %>% 
  filter(ridageyr %in% (18:19)) %>% 
     mutate(dmdeduc2 = 
            ifelse(dmdeduc3 < 9, "1", 
            ifelse(dmdeduc3 %in% (9:11), "2",
            ifelse(dmdeduc3 %in% (12:14), "3", "4")))) %>% 
  mutate(dmdeduc2 = as.numeric(dmdeduc2))
  
model_data_2 = 
  model_data %>% 
  filter(!ridageyr %in% (18:19))

model_data = bind_rows(model_data_1, model_data_2)

# recoding marital status for <20 year old participants
model_data = 
  model_data %>% 
    mutate(dmdmartl = replace_na(dmdmartl, 5))

# re-classifying variable types 
model_data = 
  model_data %>% 
  mutate(ridreth3 = as.factor(ridreth3), 
         dmdeduc2 = as.factor(dmdeduc2), 
         dmdmartl = as.factor(dmdmartl), 
         indfmin2 = as.factor(indfmin2), 
         dmdeduc2 = as.factor(dmdeduc2)) %>% 
  select(-ridreth1)
         

```


## Trying unweighted models - Phenols

```{r Benzophenone-3}

model_raw = 
  model_data %>% 
  drop_na(urxbp3) %>% 
  filter(riagendr == 2 & ridageyr > 17 & ridageyr < 45 & !indfmin2 %in% c(77,99))

#adjusted model
model_urxbp3 = lm(log(urxbp3) ~ ridageyr + factor(ridreth3) + factor(dmdeduc2) + factor(dmdmartl) + factor(indfmin2) + urxucr_x, data = model_raw) 
model_urxbp3 %>% broom::glance()
model_urxbp3 %>% broom::tidy()

#crude model
model_urxbp3_crude = lm(log(urxbp3) ~ factor(dmdeduc2) + factor(indfmin2) + urxucr_x, data = model_raw) 
model_urxbp3_crude %>% broom::glance()
model_urxbp3_crude %>% broom::tidy()

```


## Unweighted Models - Pthalates
```{r Mono-ethyl phthalate}
#adjusted model
model_urxmep = lm(log(urxmep) ~ ridageyr + factor(ridreth3) + factor(dmdeduc2) + factor(dmdmartl) + factor(indfmin2) + urxucr_x, data = model_raw) 
model_urxmep %>% broom::glance()
model_urxmep %>% broom::tidy()

#crude model
model_urxmep_crude = lm(log(urxmep) ~ factor(dmdeduc2) + factor(indfmin2) + urxucr_x, data = model_raw) 
model_urxmep_crude %>% broom::glance()
model_urxmep_crude %>% broom::tidy()

```

